Propositions as Sessions, Mechanically

### 2015-04-07 paper/formalisation.tex
Further additions/structural changes to formalisation section.

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
set of variables from being considered as the binder. In contrast, the
traditional ``exists-fresh'' approach, where the binder is only required to be
fresh within the abstraction's body, does not produce strong enough induction
principles in some cases, e.g. weakening~\cite{Aydemir:2008:EFM}. This
decision impacts upon the premises of the CP/GV typing rules (\S~\ref{sec:cp}
and \S~\ref{sec:gv}), however these changes do not alter the typing of
processes nor the semantics of the translation from GV to CP. An equivalence
between ``exists-fresh'' and cofinite quantification definitions of the
simply-typed $\lambda$-calculus is given by
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
result could be derived straightforwardly (but tediously) for the CP/GV
system.

Another aspect of the encoding is how to define terms. The intrinsic encoding
described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
approach, indexing terms by their type so as to prevent ill-formed terms from
being constructed. However, there are issues with using this encoding for
CP/GV: $(1)$ it is not immediately clear how to handle linear contexts using
the de Bruijn variable encoding, since in the intrinsic setting the
environment automatically supports weakening; $(2)$ an intrinsic encoding of
GV terms would be complicated by the need to enforce that session types occur
in certain instances (by use of a predicate); and $(3)$ well-typed terms
require extra assumptions about binder freshness which cannot reasonably be
expressed as a function type. For these reasons, an intrinsic approach does
not offer much benefit since a well-typed term relation would still need to be
defined. Therefore, I chose an extrinsic encoding for terms and define
well-typed terms as a separate inductive type.

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a simply typed $\lambda$-calculus ($\stlc$) this is
not onerous, but GV has four binding constructs, CP has six and propositions
have two. Indeed, a lot of effort was expended setting up this
infrastructure. However, I believe one of the contributions of this work is to
articulate what one needs from state of the art metatheory libraries in order
to handle session-based languages, and more generally, linear type systems.

@@

Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
tedious details relating to environment uniqueness and ordering (see below)
elided. In other words, bound variables are represented using de Brujin
indices and free variables are represented using an abstract type provided by
the Metatheory library. The rules with binders reflect this by specifying the
type of a bound variable in the constructor rather than the name. Cofinite
quantification is performed in the hypotheses to these rules by excluding a
finite set of names from opening the subprocesses where opening a process $P$
with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
$\exists$ and $\forall$, were included in the final Coq development since GV
(\S~\ref{sec:gv}) does not support polymorphism making the rules unnecessary
for the translation. Further, as noted by \citeauthor{Lee:2012}, support for
polymorphism in type variables creates an explosion of the infrastructure
requiring additional of opening and substituting definitions and lemmas. They
appear in the Figure simply for the sake of completeness.

Formalising the process calculus CP provided many challenges for the
Metatheory library since to my knowledge, the library has never been applied
to a process calculus before, and the treatment of names, binding and
substitution differ to that present in $\lambda$-calculi. For instance, in the
$\stlc$ one typically does not work up to permutation of binders, whereas in a
process calculus these kinds of structural equivalences are
standard. Additionally, substitution and opening are only defined for
names. In other words, one cannot substitute an arbitrary process term for a
name, unlike in the $\lambda$-calculus. Thus frameworks which assume
$\lambda$-calculi style operations are not immediately amenable to the CP
setting~\cite{Lee:2012}. Ordering in environments is ignored so the
development required lemmas allowing the permutation of
environments. Permutating environments is not currently supported by the
Metatheory library but some general lemmas were added as a result of this
work, and custom tactics were developed to automatically solve certain
permutation goals via simple proof search. However, there are still many cases
where manual sequences of primitive lemmas, such as appending, commutativity
and transitivity, need to be used. A lot of these cases are similar but there
would not appear to be an advantage in developing a tactic to do the
work. Doing so would create a dependency between a hard-to-debug and obscure
Ltac definition and the proofs which utilise it, becoming a burden to
change. This development philosophy treats Ltac definitions akin to macros, a
shorthand for expressing an oft-repeated sequence of tactics, and it is
doubtful much can be gained from it unless the use of Ltac definitions can be
made more transparent.

% ...leading on to the discussion of the explicit weakening...

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the various
interactions between servers and clients. contraction is not supported because
of the difficulty in specifying its principal cut elimination rule. In order
to specify the rule one requires to show the following: \todo{fix binders of
  Q}
\begin{gather*}
\tm{\nu A.(!0(A).P \mid ?z[z',z''].Q)}
\quad \becomes \quad \\
\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

where $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
sufficiently fresh name $y$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts
$\tm{Q}$ with the names in $\Gamma$.

Unfortunately, the reduction relation does not have knowledge of the
environment typing process P. This fact does not impact upon expressing
weakening since for weakening only the names (not the bindings) within the
environment are required. The paper presentation does not specify where
weakening could occur; it is tacitly assumed weakening can always be pushed
further up the derivation tree e.g. AxCut rule. a consequence of this choice
is that for the empty choice rule there may only be a single binding in the
context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}). one can compare alms with this version of cp
since it weakening rule corresponds to affine logic

% ...leading on to presentation of principal cut elimination rules...

@@

The structural equvialences and principal cut elimination rules are presented
in Figures~\ref{??} and \ref{??} respectively. These rules are expressed to
indicate some additional complexity not present in the original
presentation. For example, structural equivalences are not unique and
different process terms result depending on initial binder permutation. ${0
  <~> 1}P$ indicates the permuting of binders $0$ and $1$ in process P. The
paper presentation specifies only one possible associativity rule lending to
the specific split of the linear environments. The reduction rules were
complicated to prove due to being composed of multiple constructor instances
and differed significantly to the standard operational semantics one typically
deals with in lambda calculus formalisms.  linear environments caused a
problem with the splitting up of variables.

@@

@@

The commuting conversions are shown in Figures~\ref{??} and~\ref{??}. Notice
that permutating binders is essential to formulating the majority of these
rules.

% ...leading on to a discussion of further CP-specific issues encountered.

>>>>

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
set of variables from being considered as the binder. In contrast, the
traditional ``exists-fresh'' approach, where the binder is only required to be
fresh within the abstraction's body, does not produce strong enough induction
principles in some cases, e.g. weakening~\cite{Aydemir:2008:EFM}. This
decision impacts upon the premises of the CP/GV typing rules (\S~\ref{sec:cp}
and \S~\ref{sec:gv}), however these changes do not alter the typing of
processes nor the semantics of the translation from GV to CP. An equivalence
between ``exists-fresh'' and cofinite quantification definitions for the
simply-typed $\lambda$-calculus ($\stlc$) is given by
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
result could be derived straightforwardly (but tediously) for the CP/GV
system.

Another aspect of the encoding is how to define terms. The intrinsic encoding
described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
approach, indexing terms by their type so as to prevent ill-formed terms from
being constructed. However, there are issues with using this encoding for
CP/GV:
\begin{enumerate}
\item It is not immediately clear how to handle linear contexts using the de
  Bruijn variable encoding, since in the intrinsic setting the environment
  automatically supports weakening
\item An intrinsic encoding of GV terms would be complicated by the need to
  enforce that session types occur in certain instances (by use of a
  predicate)
\item Well-typed terms require extra assumptions about binder freshness which
  cannot reasonably be expressed as a function type.
\end{enumerate}

For these reasons, an intrinsic approach does not offer much benefit since a
well-typed term relation would still need to be defined. Therefore, I chose an
extrinsic encoding for terms and define well-typed terms as a separate
inductive type.

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a $\stlc$ this is not onerous, but GV has four
binding constructs, CP has six and propositions have two. Indeed, a lot of
effort was expended setting up this infrastructure. However, I believe one of
the contributions of this work is to articulate what one needs from state of
the art metatheory libraries in order to handle session-based languages, and
more generally, linear type systems.

@@

Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
tedious details relating to environment uniqueness and ordering (see below)
elided. In other words, bound variables are represented using de Brujin
indices and free variables are represented using an abstract type provided by
the Metatheory library. The rules with binders reflect this by specifying the
type of a bound variable in the constructor rather than a name. Cofinite
quantification is performed in the hypotheses to these rules by excluding a
finite set of names from opening the subprocesses where opening a process $P$
with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
$\exists$ and $\forall$, were not included in the final Coq development since
GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
support for polymorphism in type variables creates an explosion of the
infrastructure requiring additional of opening and substituting definitions
and lemmas. They appear in the Figure simply for the sake of completeness.

Formalising the process calculus CP provided many challenges for the
Metatheory library since to my knowledge, the library has never been applied
to a process calculus before, and the treatment of names, binding and
substitution differ to that present in $\lambda$-calculi. For instance, in the
$\stlc$ one typically does not work up to permutation of binders, whereas in a
process calculus these kinds of structural equivalences are
standard. Additionally, substitution and opening are only defined for
names. In other words, one cannot substitute an arbitrary process term for a
name, unlike in the $\lambda$-calculus. Thus frameworks which assume
$\lambda$-calculi style operations are not immediately amenable to the CP
setting~\cite{Lee:2012}.

Ordering in environments is ignored so the development required operations for
handling permutations of environments. The Metatheory library does not
currently support permutating environments, however some basic lemmas were
added as a result of this work, and custom tactics were developed to
automatically solve certain permutation goals via simple proof search. Even
so, there are still many cases where sequences of primitive lemmas, such as
those handling append, commutativity and transitivity, need to be applied
manually. A lot of these cases are similar but there would not appear to be an
advantage in developing a tactic to do the work. Doing so would create a
dependency between a hard-to-debug and obscure Ltac definition and the proofs
which utilise it, becoming a burden to change. Further, such a development
philosophy treats Ltac tactics akin to macros, a shorthand for expressing an
oft-repeated sequence of tactics, and it is doubtful much can be gained from
it unless the execution of Ltac tactics can be made more transparent.

@@

The principal cut elimination rules are presented in Figure~\ref{??}. The
reduction rules were complicated to prove due to being composed of multiple
constructor instances and differ significantly to the standard operational
semantics one typically deals with in $\lambda$-calculus formalisms. As a
result, these well-typed reductions were not proved ``in-place'', instead each
rule is proved as a separate lemma. Most proofs followed a pattern but it was
difficult to automate satisfactorily for the reasons stated in the previous
paragraph. However, some tasks could be delegated to a custom tactic, for
example instantiating cofinitely quantified hypotheses with a newly introduced
fresh variable.

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the various
interactions between servers and clients ($\beta_{!?}$ and $\beta_{!W}$). The
paper presentation does not specify where weakening could occur; it is tacitly
assumed weakening can always be pushed further up the derivation tree to a
base rule e.g. $\Ax\Cut$ rule. Contraction is not supported in the
formalisation due to the difficulty in specifying its principal cut
elimination rule. In order to specify the rule one requires to show the
following:
\begin{gather*}
\tm{\nu A.(!0(A).P \mid ?z[[A]].Q)}
\quad \becomes \quad \\
\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

where $\tm{?z[[A]].Q}$ denotes contraction by $z$ of two channels of type $A$
in $Q$, $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
sufficiently fresh name $y$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts
$\tm{Q}$ with the names in $\Gamma$.

Unfortunately, the reduction relation does not have knowledge of the
environment, $\Gamma$, typing process $\tm{P}$. This fact does not impact upon
expressing weakening because only the names (not the types in the bindings)
within the environment are required for weakening. one can compare alms with this version of cp
since it weakening rule corresponds to affine logic

In the empty choice rule there may only be a single binding in the context
(cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}). While this diverges with the presentation of
linear logic it became necessary to do this because otherwise there would be
no relationship between the free variables within a process and its
environment. no there is a clear iff relationship which permits reduction
rules to apply such as i think weakening but it should be any rule which takes
some fvs of proc and infers something about environments from that...otherwise
what it is not at all clear how to encode the reductions in a relation, the
subprocesses environments are created from destructing the environment

@@

\input{cpstructeq}

@@

The structural equivalences and commuting conversions are shown in
Figure~\ref{??} and Figures~\ref{??} and~\ref{??}, respectively. These rules
are expressed to indicate some additional complexity not present in the
original presentation. For example, structural equivalences are not unique and
different process terms result depending on initial binder
permutation. $\permb{0}{1}$ indicates the permuting of binders $0$ and $1$ in
process $\tm{P}$. The paper presentation specifies only one possible
associativity rule lending to the specific split of the linear environments.
Notice that permutating binders is essential to formulating the majority of
these rules. Analysing permutations containing singleton bindings was
necessary for the commuting conversion rules. Singleton bindings can occur
frequently during a proof and it is sometimes necessary to analyse a
permutation of the form P++x~A ~~ Q++R to determine whether x occurs in Q or
R. the rule for associativity requires this and so do quite a few reduction
rules. since this involves some element of searching the context followed by
destructing hypotheses it could be effectively automated in permutations in
subprocesses which use a binder of the parent. the rule for associativity w
linear environments caused a problem with the splitting up of variables.
